Disk Winds Matter: Simulating the Optical Spectra of Cataclysmic Variables

### 2014-09-02 draft.tex
implemented some of CK changes

\begin{abstract}
High-State non-magnetic Cataclysmic Variables (CVs) 
exhibit strong Hydrogen \& Helium recombination lines in their optical 
spectra. These lines can be single peaked, and may be formed 
in the same bipolar outflow that is responsible for the
blue-shifted absorption lines seen in the Ultraviolet (UV).
Here we present results obtained by incorporating a `macro-atom' treatment into
our Monte Carlo radiative transfer code, which was
originally used to synthesize the UV spectrum of a disk-dominated CV 
with a simple biconical wind. Using an identical model, we show that the wind leaves a noticeable imprint 
on the optical spectrum, producing He II 4686 and H${\alpha}$ as well as
enhanced emission in the Ly-$\alpha$ and He II $1640$ UV lines. We find that 
a boundary layer is not essential to produce significant recombination to 
He II, and if one is present we favour temperatures of $<10^5$K to avoid
overionizing the UV resonance line species such as CIV. We also conduct Zanstra method
calculations and estimate that one would infer a boundary layer temperature of around 80,000K
from our benchmark calculation, despite it not being present in the model itself. This suggests that previous calculations
done using this method are strongly dependent on both the assumed geometry and disk
spectrum.
We also note that
the presence of a boundary layer becomes more noticeable in lower accretion rate systems.
In addition, the improved treatment of recombination means that the bound-free continuum
emission is enhanced, albeit by a modest amount.
A short exploration of a 3-variable kinematic parameter space leads to models with higher densities 
that are able to produce single-peaked line emission, and also generate sufficient recombination emission
to fill in the Balmer jump photoabsorption edge.
Finally, we present a synthetic spectrum of a next generation model in and out of eclipse,
and identify the strengths and shortcomings of the model in the context of the high inclination
Nova-like variable RW Tri.
\end{abstract}

@@

Outflows are
close to ubiquitous  in accreting systems across many orders of magnitude in mass,
from young stellar objects \citep{bunn1995} right up 
to quasi-stellar objects \citep[QSOs;][]{ganguly2008}. Outflows can take the form of jets, such as those
seen in X-ray binaries \citep{fender2006} or active galactic nuclei \citep[AGN;][]{marscher2006}, but can also manifest as `winds' emanating from
the accretion disk.
Understanding the impact of accretion disk winds on spectral features has far-reaching implications. 
They provide a feedback mechanism to limit the growth of black holes in AGN \citep{silkrees1998},
and their ever-present nature implies a deep connection with the accretion process. This connection
is important to understand in order to probe the universality of accretion \citep{mchardy2006, SKUK2012}, 
and to determine the true extent to which
accretion disks conform to the `$\alpha$-disk' model proposed by 
\cite{shakurasunyaev1973}. 
Despite this, in many types of systems the impact of winds on observational
features is somewhat unknown, and many theoretical questions relating to their origin 
and driving mechanism remain unanswered.

Cataclysmic variables (CVs) are systems in which a white dwarf accretes matter from a donor
star via Roche-lobe overflow. In non-magnetic, disk-dominated systems this accretion
is mediated by an accretion disk which forms around the white dwarf. 
Nova-like (NL) variables are a particular subclass of object in which the accretion disk
is in a permanent state of relatively high accretion rate 
($\dot{M} \sim 10^{-8}$~M$_{\odot}$~yr$^{-1}$),
making them the perfect laboratories to test the $\alpha$-disk model.
For over three decades, it has been known that winds emanating from the accretion disk
are important in shaping the ultraviolet (UV) spectra of high-state CVs (Heap 1978), 
the most spectacular evidence being the P-Cygni like profiles of resonance lines such as 
C${\textsc {iv}}$ (see e.g. Cordova \& Mason 1982\nocite{cordova1982}).

While the effect of the wind on the UV is at least partially clear, it may also have more subtle 
effects across the spectral range. It has been proposed that an accretion disk wind
can cause single-peaked emission line profiles in the optical (Murray \& Chiang 1996), 
such as those seen in Nova-like variables and SW Sex stars (Honeycutt, Schlegel,
\& Kaitchuck 1986; Dhillon \& Rutten 1995). It is particularly telling that single-peaked 
lines can be seen even in eclipse, suggesting 
that a spatially extensive wind may be the source of emission (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence of the influence of winds in the optical can be seen in P-Cygni like signatures,
observed in lines such as \ha and He \textsc{i} $\lambda5876$ \citep{RN98}.
In addition, \cite{KLWB98} suggested
that recombination continuum emission from a disk wind or atmosphere
could be responsible for filling in the Balmer absorption edge. This provides
a potential explanation for
the absence of a pronounced `Balmer jump' in optical spectra of CVs, which has never been adequately produced 
by theoretical models or spectral synthesis.
This dense disk atmosphere or wind transition region
necessary to produce the required recombination emission
could also explain the Fe~\textsc{ii} `curtain'
seen in near-UV observations of the high inclination dwarf nova OY Car in queiscence 
\citep{horne1994}.

@@

The first attempts to model disk winds in CVs applied techniques used in stellar
wind modeling to an accretion disk geometry (Mauche \& Raymond
1987; Drew 1987; Vitello \& Shlosman 1988), and showed that mass loss rates of
$\sim10\%$ of the accretion rate
were required to produce the observed line strengths.
Using spectral synthesis codes to model resonance scattering,
Shlosman \& Vitello (1993; hereafter SV93) and Knigge, Woods \& Drew (1995; hereafter KWD95)
were both successful in reproducing the observed spectra such as the P-Cygni like
profiles in C\textsc{iv}~$\lambda1549$\AA.
Long \& Knigge (2002) expanded on the above work by presenting our code, which
solved self-consistently for thermal and ionization balance in the wind
according to the methods of Mazzali \& Lucy (1993; ML93). They successfully reproduced
UV resonance lines profiles and presented a synthesized spectrum of Z Cam
that agreed well with the observed features. Noebauer et al. (2010) expanded on this
by carrying out more detailed comparisons to RW Tri and UX UMa. In addition to these 
radiative transfer calculations, some work has also been done using Hydrodynamic wind models, 
such as that by \cite{pkdh2002}, and more recently by \cite{kusterer2014} with application to
AM Canum Venaticorum stars.

Here we present work using the same Monte Carlo radiative transfer code as LK02, \py, but incorporating additional 
`macro-atom' techniques suggested by Lucy (2002, 2003) to improve our treatment of the Hydrogen and Helium
optical lines, as well as recombination continuum emission. 
In section 2 we present the code
and provide a brief description of macro-atoms and the radiative transfer methods we employ. 
We describe the kinematics and geometry of our biconical wind model in section 3.
In section 4 we present a benchmark model, which is discussed in section 5 with particular attention 
devoted to a small-scale parameter search. In section 6 we present a `next generation'
model which better reproduces the spectral features observed in NLs, and 
finally, in section 7, we summarise our findings.

@@

\py is a Monte Carlo (MC) radiative transfer code which uses
the Sobolev approximation to treat line transfer. It operates 
in two distinct cycles. First, the ionization state, level populations
and temperature structure are calculated. This is done by
flying MC energy quanta through a pre-defined wind geometry and recording
the relevant estimators, before choosing  a new electron temperature 
in an attempt to balance heating and cooling in the plasma.
This process is repeated until the code converges on a 
solution. Second, the spectrum is synthesized by tracking
the energy packets through the wind and computing the 
emergent spectra at a number of user-specified viewing angles.

The code is capable of operating in a number of different
regimes, both in terms of the scale of the system, and the 
shape and characteristic energies of the illuminating spectrum. It is 
first described by LK02, who initially developed a biconical CV 
wind model to synthesize UV spectra. 
Higginbottom et al.\nocite{higginbottom2013} (2013; hereafter H13) presented a 
benchmark model for broad absorption line (BAL) QSOs.
As part of this effort, the ionization scheme was improved so that 
the code was capable of treating the power-law X-ray spectrum
exhibited by AGN.
This involved using a spectral model the mean intensity in a cell
computed from MC estimators for, instead of using a dilute blackbody
approximation. Compton processes
were also incorporated into the heating and cooling balance.
In addition, \cite[][SDL05]{simmacro2005} modeled
Brackett and Pfund line profiles in YSOs using \textsc{Python,} which included
implementing a `macro-atom' mode in order to correctly treat the more
complex transition matrix involved in recombination.

@@

Macro-atoms have been used to model astrophysical plasmas in 
Wolf-Rayet star winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012}
and Supernovae \citep{kasen2006, kerzendorfsim}.
SDL05 used \py itself to apply this technique to
a Hydrogen-only model. We are now able to adopt a hybrid scheme. 
Any species in which have their level populations determined by
more complex interactions than resonance transitions
can be treated as full macro-atoms. The other species are treated as so-called `simple-ions',
where they still follow the indivisible packet constraint but use the two-level atom approximation.
This allows one to preserve the fast treatment of resonance lines while providing 
an improved representation of, for example, recombination, which is important when modeling features 
such as the Balmer lines and also recombination {\sl continuum} emission. 
Recombination is expected to be an important process in the optical
spectra of CVs, and they thus make a perfect application for Lucy's 
advances in MC methods.

@@

\subsubsection{Simple-ions}

C, O, N, Ne, Na, Mg, Al, Si, S, Ar, Ca \& Fe are all included as simple-ions. 
The two-level atom is a first order approximation here, which is appropriate 
when treating resonance lines in which the lower level is at or close to the 
ground state. 
For these elements, we follow LK02 in using the ionization scheme presented by Mazzali \& Lucy (1993). 
Specifically, we use the formula
\begin{equation}
\frac{n_{j+1} n_e}{n_j} = W [\xi + W(1-\xi)]
\left(\frac{T_e}{T_R}\right)^{1/2}
\left(\frac{n_{j+1}n_e}{n_j}\right)^*_{T_R} \label{ionization}
\end{equation}
to solve for the relative ionization fractions. In this equation, the `starred' term on 
the right represents the Saha abundances, in accordance with the notation of Mihalas (1982). 
$W$ is an effective dilution factor, $\xi$ is the
fraction of recombinations going directly to the ground state, and
$T_R$ and $T_e$ are the radiation and electron temperature
respectively. LK02 provide a full description of our implementation.

The reason for using this scheme 
is twofold; First, unlike in H13, the SED can be approximated well by a dilute blackbody due to the 
absence of a power-law X-ray source, and therefore the above modification to the Saha equation
is appropriate; Second, this provides continuity from LK02 and allows us to do a like-for-like comparison,
which is especially useful in assessing the improvement in Hydrogen line emission, and any changes
that might occur in the ionization stages of the resonance line species such as CIV. 
Excitation is then calculated from LTE approximations. 
%This is because the 
% level populations are only important in working out the resonance scattering 
% opacity through the wind when modeling resonance lines.

\subsection{Atomic Data}
Atomic data is obtained from the same sources described in LK02 and updated in H13. 
The additional level and line information required to treat He as macro atom was 
obtained from \textsc{Topbase} \citep{topbase2005}. We treat H as a 20-level atom as in SDL05, where 
each level is defined by the principal quantum number n, which is appropriate due to
the degeneracy of each level in the H atom. He~\textsc{II} is treated in much the same way,
with 10 levels, but He~\textsc{I} has larger energy differentials between different l subshells
and triplet and singlet states. Thus, we still include levels up to $n=10$ but explicitly 
treat the $l$ and $s$ sub-orbitals as distinct levels instead of assuming they are perfectly `l-mixed',
which also allows the model to produce singlet and 
triplet lines, as seen in optical CV spectra.

@@

\py has previously been tested against a number of radiative transfer and
photoionization codes. Comparisons of ionization balance with \cld \citep{cloudy2013} 
can be found in LK02 and H13 showing excellent agreement. We have also conducted comparisons
of ionization and spectral synthesis with the supernova code \textsc{Tardis.} \tar is described by
\cite{kerzendorfsim}, and the spectral comparisons can be found therein.
In addition to these code validation efforts, we have also conducted tests of
the macro atom scheme in \py to verify that it does indeed solve for level populations correctly
and produces the expected level emissivities. Figure~\ref{tests} shows good agreement with 
Case B, l-mixed calculations by \cite{seaton1959}, and level
population calcaultions with \textsc{Tardis.}

@@

Here we describe some of the features of our biconical wind model. Our aim here is not
to provide a `best fit' model to the data, but rather to use as a starting point
the types of models which have already been successful in explaining UV spectra of CVs,
and examine their effect on the optical features when the treatment of line transfer
is improved.

@@

Our model follows the prescription outlined by Schlosman \& Vitello (1993; SV93). A schematic is shown
in figure~\ref{cartoon}. A smooth, biconical
disk wind rises from the accretion disk between radii $r_{min}$ and $r_{max}$. The covering fraction of the wind is 
controlled by the opening angles of the wind, $\theta_{min}$ and $\theta_{max}$, and the launch angle of
the other streamlines is given by

@@

where $r_0$ is the launch radius of the streamline.
As in LK02, we follow SV93's power law
velocity profile. The poloidal velocity, $v_l$, in our model is given by

@@

where $l$ is distance measured along a  streamline. $v_{\infty}$, the 
terminal velocity, is set to a fixed multiple of the escape velocity at the bottom
of a streamline, and $v_0$ is a constant launch velocity (set to $6$~km~s$^{-1}$).
Once the wind is launched, it accelerates with a characteristic acceleration
length $R_v$, and $\alpha$ is an exponent which controls how quickly the 
wind accelerates. Larger values of $\alpha$ cause the main region of 
acceleration to occur close to $R_v$, whereas smaller values will lead
to a fast acceleration close to the disk (see figure~\ref{acc_law}).
SV93 chose this velocity law so as to give a 
velocity gradient that varied continuously along a streamline, as well
as producing line profiles with realistic velocity spreads.
In addition to these motivations, adopting a consistent velocity law 
allows for direct comparison with models which have already been successful in 
simulating UV resonance line profiles, such as those presented by LK02 and SV93.

@@

\begin{table}
\centering
\begin{tabular}{p{3cm}p{4cm}}
Model A \\
\hline Free Parameters 	&	 Value \\ 
\hline \hline 
$M_{WD}$ 	 &	 $0.8 M_{\odot}$ \\ 
$R_{WD}$ 	 &	 $7\times10^{8}$cm\\ 
$\dot{M}_{acc}$ 	 &	 $10^{-9}~M_{\odot}yr^{-1}$\\ 
$\dot{M}_{wind}$  &	$10^{-8}~M_{\odot}yr^{-1}$\\ 
$r_{min}$ 	&	 $4 R_{WD}$\\ 
$r_{max}$ 	&	 $12 R_{WD}$ \\ 
$\theta_{min}$ 	&	 $20.0^{\circ}$ \\ 
$\theta_{max}$ 	&	 $65.0^{\circ}$ \\ 
$\gamma$ 	&	 $1$ \\ 
$v_{\infty}$ 	&	 $3v_{esc}$ \\ 
%%$R_v$ 	        &	 $7\times10^{10}$cm \\ 
$R_v$ 	        &	 $100 R_{WD}$ \\ 
$\alpha$ 	&	 $1.5$ \\
\end{tabular}
\centering
\caption{
Wind geometry parameters used in the benchmark CV model.}
\label{wind_param}
\end{table}

The principle of mass conservation means that the velocity law also 
controls the density, $\rho$, of the wind, which can be expressed in terms
of radial and vertical position $(r,z)$ as

@@

where $\phi$ here is the mass loss rate per unit area at $r_0$
and $v_z$ is the vertical velocity component. In this model, we
adopt a uniform value for $\phi$.

@@

MC quanta originate from an accretion disk and a central WD in our model. It is also possible to include
a boundary layer with user-defined temperature and emitting area. Emission from the wind is dealt with
by reprocessing of the indivisible energy packets originally produced in the disk or central WD.

\subsubsection{Disk Treatment}

@@

We hope that future simulations, coupled with a comparison to data, will 
actually address some of these problems, particularly since
the filtering effect of the wind may cause noticeable effects
on the colour of the emergent spectrum \citep{hassall}. 
Here we use a blackbody illuminating spectrum
for the ionization cycles (see figure~\ref{sed}) and to compute our MC estimators.
During the spectral synthesis stage of the simulation we use stellar atmosphere models.
This is partly because they provide the most pessimistic prediction in terms of whether recombination emission in the 
wind can fill in the Balmer jump, due to their intrinsic strong absorption beyond the Balmer edge.
For low temperatures, we use stellar atmospheres computed 
by Kurucz \citep{kurucz1991}, and for $T>50,000$~K we use models computed with 
\textsc{TLUSTY} \citep{tlusty}. The input spectra are created from these model
atmospheres using \textsc{Synspec}\footnote{http://nova.astro.umd.edu/Synspec43/synspec.html}.
%The temperature profile of our disk,
%with $\dot{M}_{acc} = 10^{-8}~M_{\odot}yr^{-1}$ can be seen in figure~\ref{t_profile}.
% more detail on why we use this hodge podge of atmosphere models?

\subsubsection{Boundary Layer}
The existence of a boundary layer in CVs arises naturally from the boundary condition
of the \cite{shakurasunyaev1973} model, since the material at the inner disk
edge must lose angular momentum in order to be deposited on
the surface of the compact object \cite[see e.g.][]{lyndenbell1974}.
Boundary layers are a possible source of
high energy photons which may be responsible for the photoionization of 
He~\textsc{ii}. 
\cite{pringlesavonije1979} predict temperatures of 
$200,000-500,000{\rm K}$ for the boundary layer,
but a number of studies have found difficulties
in producing the required ionization state of the wind
with such a hot, luminous boundary layer 
\citep[see e.g.][]{maucheraymond1987, drewverbunt1985}.
Hoare \& Drew (1991; HD91) \nocite{hoare1991}
use the Zanstra method to find boundary layer temperatures of
$\sim80,000{\rm K}$

@@

\section{A Next-Generation Model}

>>>>

\begin{abstract}
High-state non-magnetic cataclysmic variables (CVs) 
exhibit strong hydrogen \& helium recombination lines in their optical 
spectra. These lines can be single peaked, and may be formed 
in the same bipolar outflow that is responsible for the
blue-shifted absorption lines seen in the ultraviolet (UV).
Here, we present results obtained by incorporating a `macro-atom' treatment into
our Monte Carlo radiative transfer code, which was
originally used to synthesize the UV spectra of disk-dominated CVs 
with a simple biconical wind. 
Using an identical model, we show that the same wind leaves a noticeable imprint 
on the optical spectrum, producing \ha\ and \heiiopt\ as well as
enhanced emission in the Ly-$\alpha$ and \heiiuv\ UV lines. 
In addition, our improved treatment of recombination means that the bound-free continuum
emission is significant, but does not yet overcome the intrinsic absorption edge
present in the stellar atmosphere models used to treat the accretion disk.
We find that a boundary layer is not essential for the production
of He II recombination lines, but also explore the effect of boundary layers
with temperatures of $80-200$kK on the ionization state of the wind
and resulting spectrum.
A short exploration of a 3-variable kinematic parameter space leads to models with higher densities 
that are able to produce single-peaked line emission, and also generate sufficient recombination emission
to fill in the Balmer jump photoabsorption edge.
Finally, we present a synthetic spectrum of a next generation model in and out of eclipse,
which shows a striking similarity with the optical spectrum of the high inclination
Nova-like variable RW Tri. The densities near the disk plane in this model
are such that \civ\ has a large collisionally excited component, in 
contrast to observations.
% We find that 
% a boundary layer is not essential to produce significant recombination to 
% He II, and if one is present we favour temperatures of $<10^5$K to avoid
% overionizing the UV resonance line species such as CIV. We also conduct Zanstra method
% calculations and estimate that one would infer a boundary layer temperature of around 80,000K
% from our benchmark calculation, despite it not being present in the model itself. This suggests that previous calculations
%done using this method are strongly dependent on both the assumed geometry and disk
%spectrum.
%We also note that
%the presence of a boundary layer becomes more noticeable in lower accretion rate systems.
\end{abstract}

@@

Outflows are ubiquitous  in accreting systems across many orders of magnitude in mass,
from young stellar objects \citep{bunn1995} right up 
to quasi-stellar objects \citep[QSOs;][]{ganguly2008}. Outflows can take the form of jets, such as those
seen in X-ray binaries \citep{fender2006} or active galactic nuclei \citep[AGN;][]{marscher2006}, 
but can also manifest as `winds' emanating from the accretion disk (REFS).
Understanding the impact of accretion disk winds on spectral features has far-reaching implications. 
They provide a feedback mechanism to limit the growth of black holes in AGN \citep{silkrees1998},
and their ever-present nature implies a deep connection with the accretion process. This connection
is important to understand in order to probe the universality of accretion \citep{mchardy2006, SKUK2012} 
and to determine the true extent to which
accretion disks conform to the `$\alpha$-disk' model proposed by 
\cite{shakurasunyaev1973}. 
Despite this, in many types of systems the impact of winds on observational
features is essentially unknown, and many theoretical questions relating to their origin 
and driving mechanism remain unanswered.

Cataclysmic variables (CVs) are systems in which a white dwarf accretes matter from a donor
star via Roche-lobe overflow. In non-magnetic, disk-dominated systems this accretion
is mediated by an accretion disk which forms around the white dwarf. 
nova-like (NL) variables are a particular subclass of objects in which the accretion disk
is in a permanent state of relatively high accretion rate 
($\dot{M} \sim 10^{-8}$~M$_{\odot}$~yr$^{-1}$),
making them the perfect laboratories to test the $\alpha$-disk model.
For over three decades, it has been known that winds emanating from the accretion disk
are important in shaping the ultraviolet (UV) spectra of high-state CVs (Heap 1978), 
the most spectacular evidence being the P-Cygni like profiles of resonance lines such as 
C${\textsc {iv}}$ (see e.g. Cordova \& Mason 1982\nocite{cordova1982}).

While the effect of the wind on the UV is at least partially clear, it may also have more subtle 
effects across the spectral range. It has been proposed that an accretion disk wind
can cause single-peaked emission line profiles in the optical (Murray \& Chiang 1996), 
such as those seen in Nova-like variables and, especially, the SW Sex stars (Honeycutt, Schlegel,
\& Kaitchuck 1986; Dhillon \& Rutten 1995). It is particularly telling that single-peaked 
lines in high-inclination systems can be seen even in eclipse, suggesting 
that a spatially extensive wind may be the source of emission (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence of the influence of winds in the optical can occasionally 
be seen in P-Cygni like signatures, observed in lines such as \ha\ 
and He \textsc{i} $\lambda5876$ \citep{RN98}.
In addition, \cite{KLWB98} suggested
that recombination continuum emission from a disk wind or atmosphere
could be responsible for filling in the Balmer absorption edge. This provides
a potential explanation for
the absence of a pronounced `Balmer jump' in optical spectra of CVs, which has never been adequately produced 
by theoretical models or spectral synthesis.
This dense disk atmosphere or wind transition region
necessary to produce the required recombination emission
could also explain the Fe~\textsc{ii} `curtain'
seen in near-UV observations of the high inclination dwarf nova OY Car in quiescence 
\citep{horne1994}.

@@

The first attempts to model disk winds in CVs applied techniques used in stellar
wind modeling to an accretion disk geometry (Mauche \& Raymond
1987; Drew 1987; Vitello \& Shlosman 1988). These efforts showed that mass loss rates of
$\sim10\%$ of the accretion rate
were required to produce the observed line strengths.
Using spectral synthesis codes to model resonance scattering,
Shlosman \& Vitello (1993; hereafter SV93) and Knigge, Woods \& Drew (1995; hereafter KWD95)
were both successful in reproducing the observed spectra such as the P-Cygni like
profiles in C\textsc{iv}~$\lambda1549$\AA.
Long \& Knigge (2002; LK02) presented a code that
solved self-consistently for thermal and ionization balance in the wind
according to the methods of Mazzali \& Lucy (1993; ML93). They successfully reproduced
UV resonance lines profiles and presented a synthesized spectrum of Z Cam
that agreed well with the observed features. Noebauer et al. (2010) used the same code 
to carry out more detailed comparisons to RW Tri and UX UMa. In addition to these 
radiative transfer calculations, some work has also been done using hydrodynamic wind models, 
such as that by \cite{pkdh2002}, and more recently by \cite{kusterer2014} with application to
AM Canum Venaticorum stars.

Here, we present work using the same Monte Carlo radiative transfer code as LK02, 
\py, but incorporating additional 
`macro-atom' techniques suggested by Lucy (2002, 2003). 
This improves our treatment of the Hydrogen and Helium
optical lines, as well as recombination continuum emission. 
Our goal is to investigate the effect of existing wind models
on the optical spectra of CVs.
In section 2 we present the code
and provide a brief description of macro-atoms and the radiative transfer methods we employ. 
We describe the kinematics and geometry of our biconical wind model in section 3.
In section 4, we present a benchmark model, which is discussed in section 5 with particular attention 
devoted to a small-scale parameter search. In section 6, we present a `next generation'
model which better reproduces the spectral features observed in NLs, and, 
finally, in section 7, we summarise our findings.

@@

\py is a Monte Carlo (MC) radiative transfer code which uses
the Sobolev approximation to treat line transfer. It operates 
in two distinct cycles. First, the ionization state, level populations
and temperature structure are calculated. This is done by
flying MC energy quanta (`photons') through a pre-defined wind geometry and recording
the relevant estimators, before choosing  a new electron temperature 
in an attempt to balance heating and cooling in the plasma.
This process is repeated until the code converges on a 
solution. Second, the spectrum is synthesized by tracking
the energy packets through the wind and computing the 
emergent spectra at a number of user-specified viewing angles.

The code is capable of operating in a number of different
regimes, both in terms of the scale of the system, and the 
shape and characteristic energies of the illuminating spectrum. 
It was originally described by LK02, who developed a biconical CV 
wind model to synthesize UV spectra. 
Higginbottom et al.\nocite{higginbottom2013} (2013; hereafter H13) used the same code to
present a benchmark model for broad absorption line (BAL) QSOs.
As part of this effort, the ionization scheme was improved so that 
the code is now capable of treating arbritrary SEDs.
%the power-law X-ray spectrum exhibited by AGN.
% This involved using a spectral model the mean intensity in a cell
% computed from MC estimators for, instead of using a dilute blackbody
% approximation. Compton processes
% were also incorporated into the heating and cooling balance.
In addition, \cite[][SDL05]{simmacro2005} modeled
Brackett and Pfund line profiles in YSOs using \textsc{Python,} which included
implementing a `macro-atom' mode in order to correctly treat the more
complex transition matrix involved in recombination.

@@

Macro-atoms have been used to model astrophysical plasmas in 
Wolf-Rayet star winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012}
and supernovae \citep{kasen2006, kerzendorfsim}.
SDL05 used \py itself to apply this technique to
a Hydrogen-only model. We are now able to adopt a hybrid scheme. 
Any species in which have their level populations determined by
more complex interactions than resonance transitions
can be treated as full macro-atoms. The other species are treated as so-called `simple-ions',
where they still follow the indivisible packet constraint but use the two-level atom approximation.
This allows one to preserve the fast treatment of resonance lines while providing 
an improved representation of, for example, recombination, which is important when modeling features 
such as the Balmer lines and also recombination {\sl continuum} emission. 
Recombination is expected to be an important process in the optical
spectra of CVs, and they thus make a perfect application for Lucy's 
advances in MC methods.

@@

\subsubsection{Two-level atoms}

C, O, N, Ne, Na, Mg, Al, Si, S, Ar, Ca \& Fe are all included as two-level atoms. 
This is a first order approximation, which is appropriate 
when treating resonance lines in which the lower level is at or close to the 
ground state. 
For these elements, we follow LK02 in using the ionization scheme presented by Mazzali \& Lucy (1993). 
Specifically, we use the formula
\begin{equation}
\frac{n_{j+1} n_e}{n_j} = W [\xi + W(1-\xi)]
\left(\frac{T_e}{T_R}\right)^{1/2}
\left(\frac{n_{j+1}n_e}{n_j}\right)^*_{T_R} \label{ionization}
\end{equation}
to solve for the relative ionization fractions. In this equation, the `starred' term on 
the right represents the Saha abundances, in accordance with the notation of Mihalas (1982). 
$W$ is an effective dilution factor, $\xi$ is the
fraction of recombinations going directly to the ground state, and
$T_R$ and $T_e$ are the radiation and electron temperature
respectively. LK02 provide a full description of our implementation.

The reason for using this scheme 
is twofold. First, unlike in H13, the SED can be approximated well by a dilute blackbody due to the 
absence of a power-law X-ray source, and therefore the above modification to the Saha equation
is appropriate; Second, this provides continuity from LK02 and allows us to do a like-for-like comparison.
This is especially useful in assessing the improvement in Hydrogen line emission, and any changes
that might occur in the ionization stages of the resonance line species such as CIV. 
Excitation is then calculated in LTE. 
%This is because the 
% level populations are only important in working out the resonance scattering 
% opacity through the wind when modeling resonance lines.

\subsection{Atomic Data}
Atomic data is obtained from the same sources described in LK02 and updated in H13. 
The additional level and line information required to treat He as a macro-atom was 
obtained from \textsc{Topbase} \citep{topbase2005}. We treat H as a 20-level atom as in SDL05, where 
each level is defined by the principal quantum number, $n$, which is appropriate due to
the degeneracy of each level in the H atom. He~\textsc{II} is treated in much the same way,
but with 10 levels. However, He~\textsc{I} has larger energy differentials between different l subshells
and triplet and singlet states. Thus, we still include levels up to $n=10$ but explicitly 
treat the $l$ and $s$ sub-orbitals as distinct levels instead of assuming they are perfectly `l-mixed',
This also allows the model to produce singlet and 
triplet lines, which are seen in optical CV spectra \citep[e.g.][]{dhillon1996}.

@@

\py has previously been tested against a number of radiative transfer and
photoionization codes. Comparisons of ionization balance with \cld \citep{cloudy2013} 
can be found in LK02 and H13 and show excellent agreement. We have also conducted comparisons
of ionization and spectral synthesis with the supernova code \textsc{Tardis.} \tar is described by
\cite{kerzendorfsim}, and the spectral comparisons can be found therein.
In addition to these code validation efforts, we have also conducted tests of
the macro atom scheme in \py to verify that it does indeed solve for level populations correctly
and produces the expected level emissivities. Figure~\ref{tests} shows good agreement with 
Case B, l-mixed calculations by \cite{seaton1959}, and level
population calcaultions with \textsc{Tardis.}

@@

Here we describe some of the features of our biconical wind model. Our aim here is not
to provide a `best fit' model to match any particular dataset.
Rather, we take as a starting point
the types of models which have already been successful in explaining UV spectra of CVs.
We can then examine the contribution of the wind to 
the optical features when the treatment of line transfer
is improved.

@@

Our model follows the prescription outlined by Schlosman \& Vitello (1993; SV93). A schematic is shown
in figure~\ref{cartoon}. A smooth, biconical
disk wind rises from the accretion disk between radii $r_{min}$ and $r_{max}$. The covering fraction of the wind is 
also controlled by the opening angles of the wind, $\theta_{min}$ and $\theta_{max}$, and the launch angle of
the other streamlines is given by

@@

where $r_0$ is the launch radius of the streamline.
%As in LK02, we follow SV93's power lawvelocity profile. 
The poloidal velocity, $v_l$, in the SV93 model is given by

@@

where $l$ is distance measured along a  streamline. The terminal velocity, 
$v_{\infty}$, is set to a fixed multiple of the escape velocity at the bottom
of a streamline, and $v_0$ is a constant launch velocity (set to $6$~km~s$^{-1}$).
Once the wind is launched, it accelerates with a characteristic acceleration
length $R_v$, and $\alpha$ is an exponent which controls how quickly the 
wind accelerates. Larger values of $\alpha$ cause the main region of 
acceleration to occur close to $R_v$, whereas smaller values will lead
to a fast acceleration close to the disk (see figure~\ref{acc_law}).
SV93 chose this velocity law so as to give a 
velocity gradient that varied continuously along a streamline, as well
as producing line profiles with realistic velocity spreads.
In addition to these motivations, adopting a consistent velocity law 
allows for direct comparison with models which have already been successful in 
simulating UV resonance line profiles, such as those presented by LK02 and SV93.

@@

\begin{table}
\centering
\begin{tabular}{p{3cm}p{4cm}}
Model A \\
\hline Free Parameters 	&	 Value \\ 
\hline \hline 
$M_{WD}$ 	 &	 $0.8 M_{\odot}$ \\ 
$R_{WD}$ 	 &	 $7\times10^{8}$cm\\ 
$\dot{M}_{acc}$ 	 &	 $10^{-9}~M_{\odot}yr^{-1}$\\ 
$\dot{M}_{wind}$  &	fi\\ 
$r_{min}$ 	&	 $4 R_{WD}$\\ 
$r_{max}$ 	&	 $12 R_{WD}$ \\ 
$\theta_{min}$ 	&	 $20.0^{\circ}$ \\ 
$\theta_{max}$ 	&	 $65.0^{\circ}$ \\ 
$\gamma$ 	&	 $1$ \\ 
$v_{\infty}$ 	&	 $3v_{esc}$ \\ 
%%$R_v$ 	        &	 $7\times10^{10}$cm \\ 
$R_v$ 	        &	 $100 R_{WD}$ \\ 
$\alpha$ 	&	 $1.5$ \\
\end{tabular}
\centering
\caption{
Wind geometry parameters used in the benchmark CV model.}
\label{wind_param}
\end{table}

%The principle of mass conservation 
The mass continuity equation means that the velocity law also 
controls the density, $\rho$, of the wind, which can be expressed in terms
of radial and vertical position $(r,z)$ as

@@

where $\phi$ is the mass-loss rate per unit area at $r_0$
and $v_z$ is the vertical velocity component. In our model, we
adopt a uniform value for $\phi$.

@@

MC quanta originate from the accretion disk and a central WD in our model. It is also possible to include
a boundary layer with user-defined temperature and emitting area. Emission from the wind is dealt with
by reprocessing of the indivisible energy packets originally produced in the disk or central WD.

\subsubsection{The Accretion Disk}

@@

We hope that future simulations, coupled with a comparison to data, will 
actually address some of these problems, particularly since
the filtering effect of the wind may cause noticeable effects
on the colour of the emergent spectrum \citep{hassall}. 
Here, we use a blackbody illuminating spectrum
for the ionization cycles (see figure~\ref{sed}) and to compute our MC estimators.
During the spectral synthesis stage of the simulation we use stellar atmosphere models.
This is partly because they provide the most pessimistic prediction in terms of whether recombination emission in the 
wind can fill in the Balmer jump, due to their intrinsic strong absorption beyond the Balmer edge.
For low temperatures, we use stellar atmospheres computed 
by Kurucz \citep{kurucz1991}, and for $T>50,000$~K we use models computed with 
\textsc{TLUSTY} \citep{tlusty}. The input spectra are created from these model
atmospheres using \textsc{Synspec}\footnote{http://nova.astro.umd.edu/Synspec43/synspec.html}.
%The temperature profile of our disk,
%with $\dot{M}_{acc} = 10^{-8}~M_{\odot}yr^{-1}$ can be seen in figure~\ref{t_profile}.
% more detail on why we use this hodge podge of atmosphere models?

\subsubsection{Boundary Layer}
The existence of a boundary layer in CVs arises naturally from the boundary condition
of the \cite{shakurasunyaev1973} model, since the material at the inner disk
edge must lose angular momentum in order to be deposited on
the surface of the compact object \cite[see e.g.][]{lyndenbell1974}.
Boundary layers are a possible source of
high energy photons which may be responsible for the photoionization of 
He~\textsc{ii}. 
\cite{pringlesavonije1979} predict temperatures of 
$200,000-500,000{\rm K}$ for the boundary layer,
but a number of studies have found difficulties
in producing the observed ionization state of the wind
with such a hot, luminous boundary layer 
\citep[see e.g.][]{maucheraymond1987, drewverbunt1985}.
Hoare \& Drew (1991; HD91) \nocite{hoare1991}
use the method of \cite{zanstra1929} to find upper limits on 
boundary layer temperatures of $\sim80,000{\rm K}$.

@@

\section{An Improved Optical Model }
